---
title: "Reproducing the Simulations and Empirical Analyses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the Simulations and Empirical Analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Reproduction interface

The package separates reusable statistical methods from article-specific
workflows. The following functions expose every numerical study reported in
the manuscript:

| Function | Article output |
|---|---|
| `reproduce_hdzacp_size()` | AR(1) and compound-symmetry size table |
| `reproduce_hdzacp_power()` | Student-tail and skewness power figures |
| `reproduce_hdzacp_location()` | single-change location-error figure |
| `reproduce_hdzacp_multiple()` | multiple-change count, Hausdorff, and adjusted-Rand results |
| `reproduce_hdzacp_analytic_size()` | numerical validation of the analytic null calibration |
| `reproduce_hdzacp_braincloud()` | BrainCloud global tests and single location |
| `reproduce_hdzacp_gas_sensor()` | Gas Sensor global and multiple-change analysis |
| `reproduce_hdzacp_article()` | coordinated execution of all workflows |

All functions require an explicit `output_dir` or create a clearly named
subdirectory under the supplied article directory. They return their principal
tables invisibly in addition to writing machine-readable CSV or RDS output.
Random seeds and effective settings are written alongside the results.

Use `quick = TRUE` only as an installation smoke test:

```{r smoke}
library(HDZACP)

reproduce_hdzacp_size(
  quick = TRUE,
  output_dir = "results/smoke/size"
)
reproduce_hdzacp_power(
  quick = TRUE,
  output_dir = "results/smoke/power"
)
```

Full article settings are selected by `quick = FALSE`, which is the default.
They are computationally expensive and should be run in a persistent output
directory:

```{r full}
reproduce_hdzacp_article(
  quick = FALSE,
  output_dir = "results/article"
)
```

The scripts installed in `inst/reproduce/` offer the same interface from a
terminal. `00_run_all.R --quick` performs a smoke run; omitting `--quick` uses
the full defaults.

## Global size

Under the null, (X_i\sim N_p(0,\Sigma)), independently over time. Both
dependence models use (ho=0.5):

\[
(\Sigma_{\mathrm{AR}})_{jk}=\rho^{|j-k|},\qquad
\Sigma_{\mathrm{CS}}=(1-\rho)I_p+\rho\mathbf 1_p\mathbf 1_p^\top.
\]

The grid is (n\in\{100,200\}) and (p\in\{100,200,400\}). Every cell uses
1,000 Monte Carlo replications, 199 whole-vector permutations, trimming
fraction 0.1, and nominal level 0.05. The comparison contains ZAC, ZAM, ZAS,
KDist, HDD, E-Divisive, and gSeg.

```{r size}
reproduce_hdzacp_size(
  quick = FALSE,
  output_dir = "results/article/size"
)
```

## Power under shape and tail changes

The main power study uses (n=p=200), a change after observation 100, and the
independent-component model (X_i=L_m\varepsilon_i), where
(L_mL_m^\top=\Sigma_m) for either AR(1) or compound symmetry. Before the
change all innovations are standard normal. After the change, the first (s)
innovations are standardized Student variables and the remaining innovations
stay normal. Thus the mean and covariance do not change.

The sparsity grid is

\[
s\in\{1,2,5,10,20,30,40,50,60,80,120,200\}.
\]

For AR(1), the degrees of freedom are

\[
\nu_{\mathrm{AR}}(s)=\min\{30,2.0001+0.1\log_2 s\}.
\]

For compound symmetry, let

\[
w(s)=\begin{cases}
0,&s\le20,\\
\log_2(s/20)/\log_2(50/20),&20<s<50,\\
1,&s\ge50,
\end{cases}
\]

and use

\[
\nu_{\mathrm{CS}}(s)=\min\left\{30,2.0001+
0.1\{1-w(s)\}\log_2s+0.05w(s)\log_2(p/s)\right\}.
\]

Two additional AR(1) experiments replace the affected Student innovations by
(\operatorname{Exp}(1)-1) and by
(\{\operatorname{Gamma}(4,1)-4\}/2). Each configuration uses 1,000
replications and 199 permutations.

```{r power}
reproduce_hdzacp_power(
  quick = FALSE,
  output_dir = "results/article/power"
)
```

## Single-change localization

The location experiment uses the same AR(1) and compound-symmetry Student-tail
models. Every method returns one maximizing split, and the article reports
(|\widehat\tau-100|/200). ZAC inherits the location of the component with the
smaller permutation p-value. The full experiment again uses 1,000 replications
and 199 permutations for every configuration.

```{r location}
reproduce_hdzacp_location(
  quick = FALSE,
  output_dir = "results/article/location"
)
```

## Multiple changes

The multiple-change design uses (n=p=200), AR(1) dependence with
(ho=0.5), and true changes at 60 and 140. The first (s) innovations are
normal, standardized Student, and normal in the three consecutive regimes; all
remaining innovations stay normal. The degrees of freedom follow the AR(1)
rule above.

SUM-WBS, MAX-WBS, and Cauchy-adaptive WBS use a shared set of 50 random
intervals plus the full interval, minimum random-interval length 40, local
trimming fraction 0.1, 99 local permutations, and unadjusted 5% stopping. KDist
uses its native WBS implementation, HDD and gSeg use the article's recursive
implementations, and E-Divisive uses its native divisive recursion. Each value
of (s) has 100 Monte Carlo replications.

The three criteria are the probability of estimating exactly two changes,
scaled Hausdorff loss, and the adjusted Rand index of the induced segmentation.
For an empty estimate, the Hausdorff loss is set to (n/L_{\max}), where
(L_{\max}) is the longest true segment.

```{r multiple}
reproduce_hdzacp_multiple(
  quick = FALSE,
  output_dir = "results/article/multiple"
)
```

## Analytic null calibration

The analytic study is separate from the permutation-based method comparison.
It validates the corrected SUM, MAX, and Cauchy null laws over the article's
specified ((n,p)) pairs and dependence designs, and records size, distribution
diagnostics, numerical sensitivity checks, and calibration constants.

```{r analytic-size}
reproduce_hdzacp_analytic_size(
  quick = FALSE,
  output_dir = "results/article/analytic_size"
)
```

## BrainCloud analysis

No expression data are distributed with `HDZACP`. Download the normalized
Series Matrix from the official GEO record:

- [GSE30272 landing page](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE30272)
- [GSE30272 Series Matrix](https://ftp.ncbi.nlm.nih.gov/geo/series/GSE30nnn/GSE30272/matrix/GSE30272_series_matrix.txt.gz)

The article keeps subjects aged at least 20 years, adjusts each probe for an
intercept and the two supplied surrogate variables without adjusting for age,
removes nonfinite and zero-variance probes, retains the 2,000 largest residual
variances, applies featurewise median/MAD standardization, and orders subjects
by age. This yields (n=148), (p=2000). All seven methods use the same matrix,
trimming fraction 0.1, and 999 permutations.

```{r braincloud}
reproduce_hdzacp_braincloud(
  data_file = "data/GSE30272_series_matrix.txt.gz",
  output_dir = "results/article/braincloud"
)
```

The estimated age corresponding to split (k) is the midpoint between the
observed ages at rows (k) and (k+1). The result is an age-ordered
cross-sectional transition and must not be interpreted as a longitudinal
within-person or causal aging threshold.

## Gas Sensor Array Drift analysis

No sensor data are distributed with `HDZACP`. Obtain the archive from UCI:

- [UCI landing page](https://archive.ics.uci.edu/dataset/270/gas%2Bsensor%2Barray%2Bdrift%2Bdataset%2Bat%2Bdifferent%2Bconcentrations)
- [Official ZIP archive](https://archive.ics.uci.edu/static/public/270/gas+sensor+array+drift+dataset+at+different+concentrations.zip)
- [Dataset DOI](https://doi.org/10.24432/C5MK6M)

The article retains class-1 ethanol measurements and 128 sensor features.
Featurewise preprocessing consists of robust standardization, a natural-spline
location regression on log concentration with four degrees of freedom, an
analogous scale regression, and final median/MAD standardization. Up to 50
measurements are sampled without replacement from each of the ten batches. The
retained counts are (50,50,50,50,28,50,50,30,50,50), giving (n=458).
Rows are randomized within batch using the fixed article seed, while batch
order is preserved.

Global tests use 999 permutations. The proposed multiple-change methods use
100 WBS intervals and 199 local permutations. The nine adjacent-batch
boundaries are reference transitions, not change points imposed on the
algorithms.

```{r gas}
reproduce_hdzacp_gas_sensor(
  data_zip = "data/gas_sensor_array_drift.zip",
  output_dir = "results/article/gas_sensor"
)
```

The interpretation is descriptive. Exact acquisition times within a batch are
unavailable, and concentration support differs across batches. The location
and scale adjustment reduces the concentration effect but cannot remove
confounding where batch-specific concentration supports do not overlap.

## Optional comparison packages

The full seven-method workflows need three CRAN packages and two GitHub
packages in addition to the core package:

```{r install, eval=FALSE}
install.packages(c("ade4", "ecp", "gSeg", "remotes"))
remotes::install_github("zhangxiany-tamu/KDist")
remotes::install_github("rezadrikvandi/HDDchangepoint")
```

The output metadata records the package versions and the precise wrapper used.
This is important because KDist and HDDchangepoint are development packages,
and because permutation conventions differ across methods.
