---
title: "Methods, Calibration, and Change-Point Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods, Calibration, and Change-Point Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Statistical target

Let the rows of an (n\times p) matrix be an ordered sequence
(X_1,\ldots,X_n). The procedures in `HDZACP` test whether at least one
coordinatewise marginal distribution changes along the sequence and estimate
the associated split locations. A change in location, scale, tail thickness,
skewness, modality, or another marginal feature is within scope. A pure copula
change for which every marginal law stays fixed is not the target.

At each admissible split (k), the package applies a likelihood-ratio-type
two-sample rank statistic to every coordinate. Its exact finite-sample null
mean and variance depend on (n) and (k), but not on an unknown continuous
marginal law. Standardized coordinatewise deficits are then aggregated in two
ways. ZAS sums evidence over coordinates and is suited to dense changes; ZAM
takes the coordinatewise maximum and is suited to sparse changes. ZAC combines
the two pieces of evidence and adapts to unknown sparsity.

Candidate splits are restricted by the trimming fraction `eta`. With
`eta = 0.1`, the package scans from `ceiling(0.1 * n)` through
`n - ceiling(0.1 * n)`. Returned estimates are split-after indices, so an
estimate (k) separates rows (1,\ldots,k) from rows (k+1,\ldots,n).

## Single-change interface

The primary function evaluates any subset of the three procedures in one call:

```{r single}
library(HDZACP)

set.seed(7)
n <- 180
p <- 120
x <- matrix(rnorm(n * p), n, p)
x[91:n, 1:8] <- matrix(rt((n - 90) * 8, df = 3) / sqrt(3), n - 90, 8)

fit <- hdzacp_test(
  x,
  methods = c("ZAC", "ZAM", "ZAS"),
  calibration = "permutation",
  eta = 0.1,
  alpha = 0.05,
  permutations = 199,
  seed = 20260923
)

fit$results
fit$components
```

The result's `results` component contains one row per requested method, with
the statistic, p-value, rejection decision, estimated split, estimated split
fraction, selected component, and calibration. The `components` entry records
the ZAS and ZAM statistics, p-values, and locations used by ZAC. The convenience
functions `zac_test()`, `zam_test()`, and `zas_test()` forward their arguments to
`hdzacp_test()`.

For ZAC, the estimated location is inherited from the component with the
smaller p-value; an exact tie selects the ZAS location. This rule is part of the
estimator and should be reported when ZAC is used for localization.

## Whole-vector permutation calibration

For the default `calibration = "permutation"`, an entire row of (X) is moved
as one unit. This preserves arbitrary contemporaneous dependence among the
coordinates. Under the null, exact permutation validity requires exchangeable
observation vectors. Serial dependence or a deterministic time trend under the
null invalidates unrestricted row permutation unless a dependence-respecting
calibration is supplied.

The observed ordering and `B = permutations` random orderings form a symmetric
orbit of size (B+1). ZAS and ZAM p-values are ranks within this orbit. For
each orbit member, ZAC combines the orbit ranks of both components and ranks the
combined score over the same orbit. This construction includes the observed
ordering symmetrically and has minimum attainable p-value (1/(B+1)).

Consequently, the permutation count must be chosen with the target level in
mind. For a 5% test, 199 permutations give resolution 0.005; 999 permutations
give resolution 0.001. Inside WBS, a Bonferroni threshold smaller than
(1/(B+1)) cannot reject any local interval and should trigger a diagnostic.

## Analytic calibration

```{r analytic}
fit_analytic <- hdzacp_test(
  x,
  methods = c("ZAC", "ZAM", "ZAS"),
  calibration = "asymptotic",
  eta = 0.1
)
fit_analytic$results
```

The analytic option implements the null laws established in the article. ZAS
uses its Gaussian-process supremum limit, while ZAM uses the appropriate
extreme-value normalization when the dimension-growth regime is supported by
the implemented constants. The current analytic MAX calibration is restricted
to the subcritical branch. Unsupported critical or boundary inputs cause an
informative failure rather than substitution of an unrelated limit. The
analytic Cauchy calibration uses the established asymptotic independence of
the two component statistics.

Permutation and analytic calibration answer the same null question but rely on
different assumptions. Permutation is the routine default when row
exchangeability is plausible. Analytic calibration is useful for studying the
theory and for settings that satisfy the article's Gaussian-copula and
dimension-growth assumptions.

## Ties and input checks

The rank theory is stated for continuous marginal distributions. The default
`ties = "error"` avoids silently applying continuous-rank formulas to tied
data. With permutation calibration, `ties = "random"` and a fixed `tie_seed`
provide reproducible random tie breaking, applied consistently to the observed
and permuted orderings. Random tie breaking is not accepted for analytic
calibration.

Input must be a finite numeric matrix with observations in rows. A data frame
should be converted only after nonnumeric identifiers and ordering variables
have been removed. Preprocessing must not use information from a suspected
post-change regime in a way that artificially aligns or separates the two
sides of a candidate split.

## Multiple changes

`hdzacp_wbs()` embeds the three local tests in wild binary segmentation. A
common interval collection lets ZAS, ZAM, and ZAC see the same candidate
subproblems.

```{r wbs}
wbs_fit <- hdzacp_wbs(
  x,
  methods = c("ZAC", "ZAM", "ZAS"),
  calibration = "permutation",
  eta = 0.1,
  alpha = 0.05,
  n_intervals = 100,
  min_interval = 30,
  include_full = TRUE,
  permutations = 99,
  stopping = "paper",
  seed = 20260924,
  keep_evidence = TRUE
)

wbs_fit$changepoints
wbs_fit$segments
head(wbs_fit$evidence)
```

The evidence table records interval endpoints, interval length, the three local
p-values, the ZAS and ZAM local maximizers, the adaptive maximizer, and ZAC's
selected component. The `intervals` and `settings` entries permit an exact rerun
with the same interval collection.

The available stopping choices deliberately separate different inferential
goals. The default `stopping = "bonferroni"` adjusts over the sampled interval
collection, `"paper"` applies the article's analytic stopping construction, and
`"unadjusted"` reproduces the finite-sample simulations in the article. The
article's multiple-change experiment uses 50 random intervals plus the full
interval, minimum random-interval length 40, local `eta = 0.1`, 99
permutations, and unadjusted local 5% stopping.

## External comparisons

`hdzacp_compare()` provides a common result schema but delegates the competing
methods to their authors' packages:

- KDist uses `KDist::kcpd_single()` and `KDist::kcpd_wbs()` from
  [zhangxiany-tamu/KDist](https://github.com/zhangxiany-tamu/KDist).
- HDD uses `HDDchangepoint::test_single_changepoint()` and
  `HDDchangepoint::multiple_changepoint_detection_wbs()` from
  [rezadrikvandi/HDDchangepoint](https://github.com/rezadrikvandi/HDDchangepoint).
- E-Divisive uses `ecp::e.divisive()` from the CRAN package `ecp`.
- gSeg builds an MST with `ade4::mstree()` and calls `gSeg::gseg1()` from the
  CRAN packages `ade4` and `gSeg`.

```{r comparisons}
comparison <- hdzacp_compare(x, permutations = 199, seed = 10)
comparison
```

The returned calibration field should be retained in tables. Methods with the
same nominal permutation count can still use different sequential, conditional,
or plus-one conventions. The article-reproduction functions preserve those
choices and record them rather than treating every reported p-value as if it
came from the same algorithm.

